---
title: "Introduction to testdat"
# author: "Danny Smith"
# date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to testdat}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, echo = FALSE, warning = FALSE, message = FALSE}
knitr::opts_chunk$set(collapse = T, comment = "#>", error = T)
options(tibble.print_min = 4L, tibble.print_max = 4L)
library(testdat)
```

testdat is a package designed to ease data validation, particularly for complex
data processing, inspired by software unit testing. testdat extends the strong
and flexible unit testing framework already provided by testthat with a family
of functions and reporting tools focused on checking of data frames.

Features include:

* A fully fledged test framework so you can spend more time specifying tests and
  less time running them

* A set of common methods for simply specifying data validation rules

* Repeatability of data tests (avoid unintentionally breaking your dataset!)

* Data-focused reporting of test results

## Getting started

As an extension of testthat, testdat uses the same basic testing framework.
Before using testdat (and reading this documentation), make sure you're familiar
with the introduction to testthat in
[R packages](http://r-pkgs.had.co.nz/tests.html).

The main addition in testdat is a set of expectations designed for testing data
frames and a mechanism to attach a data frame to a testing context.

### Data Expectations

Data expectations fall into a few main classes.

Generic

:   Man page: [generic-expectations](../help/generic-expectations)

    Generic expectations allow for testing of a dataset using an arbitrary
    function. The function provided should take a single vector as its first
    argument and return a logical vector showing whether each element has passed
    or failed. Additional arguments to the checking function can be passed as a
    list using the `args` argument.
  
    testdat includes a set of useful checking functions, a list can be found on
    the [check_generic](../help/check_generic) man page.
  
Conditional

:   Man page: [conditional-expectations](../help/conditional-expectations)

    Conditional expectations check for the co-existence of multiple conditions.

Value

:   Man page: [value-expectations](../help/value-expectations)

    Value expectations test variable values for valid data. Tests include
    explicit value checks, uniqueness and others.

Dataset Comparison

:   Man page: [datacomp-expectations](../help/datacomp-expectations)

    Dataset comparison expectations test for consistency between datasets, for
    example ensuring similar frequencies between similar variables in different
    datasets.

### Data Context

Each set of tests in testthat has a `context()`. In testdat, each context has an
associated dataset, specified with a call to `context_data()`.

```{r}
context("Example")
context_data(iris)

test_that("Variable format checks", {
  expect_regex(Species, "^[a-z]+$")
})

test_that("Versicolor has sepal length greater than 5 - will fail", {
  expect_cond(Species %in% "versicolor", Sepal.Length >= 5)
})
```

A call to `context_data()` sets the value of the "global test data set", which
is used as the default for the `data` argument of the various data expectations.
`set_testdata()` can also be used to modify the global test data outside of a
testing context.

```{r}
set_testdata(mtcars)
identical(get_testdata(), mtcars)

args(expect_cond)
```
